% Unitary Model of Consciousness — Mathematical Theorems
% Author: Olkhovoy

\documentclass[12pt]{article}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{hyperref}
\usepackage{geometry}
\geometry{margin=1in}

\title{Formal Theorems in the Unitary Model of Consciousness}
\author{A. Olkhovoy}
\date{\today}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

\begin{document}

\maketitle

\section*{Theorem: Rotational Unitary Consciousness}

\begin{theorem}[Rotational Unitary Consciousness]
Let $A = \{a_1, \dots, a_N\}$ be a finite set of agents. Consider an infinite sequence of discrete iterations $t = 1, 2, \dots$. In each iteration, exactly one agent $i_t \in A$ is the Active Agent, and all others are Echoes $j \in A \setminus \{i_t\}$. Suppose the selection of the active agent follows a fair rotational rule: each agent appears infinitely often and asymptotically equally.

Let $M_t$ be the world model at iteration $t$, updated by the Active Agent via predictive error minimization. Then:

\begin{enumerate}
    \item Each agent $a \in A$ is active infinitely often, and
    \[ \lim_{T \to \infty} \frac{1}{T} \left|\{ t \leq T : i_t = a \} \right| = \frac{1}{N}. \]

    \item The model $M_t$ converges under predictive error minimization:
    \[ \| O_t - \hat{O}_t \| \to 0, \quad \text{as } t \to \infty. \]

    \item The update process is Turing-computable.
\end{enumerate}

Thus, the scheme is fair, convergent, and computationally grounded.
\end{theorem}

\vspace{1em}

\section*{Lemma: Ontological Limit of the Observer}

\begin{lemma}[Ontological Limit of the Observer]
Suppose the observer (Active Agent) experiences a simulated world $O_t$ generated by a world model $M_{t-1}$ under predictive processing. Then, under sufficient iteration:
\begin{enumerate}
    \item The prediction error diminishes asymptotically:
    \[ \|O_t - \hat{O}_t\| \to 0. \]
    \item The perceived world becomes increasingly stable and "normal."
    \item Chaotic events are perceived as rare outliers, not systemic disruptions.
\end{enumerate}

Consequently, if you awaken in a stable reality, you are most likely in the limit state of a long-trained simulation.
\end{lemma}

\end{document}
